---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# fastplyr

<!-- badges: start -->

[![R-CMD-check](https://github.com/NicChr/fastplyr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/NicChr/fastplyr/actions/workflows/R-CMD-check.yaml)

<!-- badges: end -->

fastplyr aims to provide a [tidyverse](https://www.tidyverse.org/learn) frontend using a [collapse](https://sebkrantz.github.io/collapse/articles/collapse_intro.html) backend. This means from a user's point of view the functions behave like the tidyverse equivalents and thus require little to no changes to existing code to convert.

fastplyr is designed to handle operations that involve larger numbers of groups and generally larger data.

## Installation

You can install the development version of fastplyr from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("NicChr/fastplyr")
```

Load packages

```{r example}
library(tidyverse)
library(fastplyr)
```

## dplyr alternatives

All tidyverse alternative functions are prefixed with 'f_'. For example, `dplyr::distinct` becomes `fastplyr::f_distinct`.
